Notes on robust solvers for systems of coupled multiscale PDEs

### 2019-08-02 solver/solver.tex
A2 done

\newcommand{\semi}[1]{\lvert{#1}\rvert}
\newcommand{\norm}[1]{\lVert{#1}\rVert}
\newcommand{\jump}[1]{\ensuremath{[\![#1]\!]} }
\newcommand{\avg}[1]{\ensuremath{\left\{\!\left\{#1\right\}\!\right\}} }
\def\ud{u_{\odot}}
\def\vd{v_{\odot}}
\def\ld{\lambda_{\odot}}
\def\md{\mu_{\odot}}
\def\lld{l_{\odot}}
\def\uf{u_{\ominus}}
\def\up{u_{\oplus}}
\def\eps{\epsilon}
\def\nn{\boldsymbol n}
\def\rr{\boldsymbol r}
\def\RR{\boldsymbol R}
\def\kk{\boldsymbol k}
\def\ss{\boldsymbol s}
\def\uu{\boldsymbol u}
\def\vv{\boldsymbol v}
\def\xx{\boldsymbol x}
\def\bu{\overline{u}}
\def\bv{\overline{v}}
\def\tu{\widetilde{u}}
\def\tv{\widetilde{v}}
\def\TT{\boldsymbol T}
\def\NN{\boldsymbol N}
\def\BB{\boldsymbol B}
\def\ttu{\widetilde{\widetilde{u}}}
\def\ttv{\widetilde{\widetilde{v}}}
\def\cv{\check{v}}
\def\mesh{{\cal T}^h}
\def\ball{{\cal B}}
\def\R{\mathbb{R}}
\def\D{\mathcal{D}}
\def\DD{\partial\mathcal{D}}
\def\trace{{\mathcal{T}_\Gamma}}
\def\mtrace{{\overline{\mathcal{T}}_\Lambda}}
\def\ext{\mathcal{E}_\Gamma}
\def\ide{\mathcal{I}}
\def\ii{\hat{\imath}}
\newcommand{\avrd}[1]{\overline{\overline{#1}}}
\newcommand{\avrc}[1]{\overline{#1}}
\newcommand{\refe}[1]{{#1}_{\mathrm{ref}}}

@@

  % Boundaries
  \draw[red!60!white] (0, 0) -- (0, 4);
  \draw[red!60!white] (4, 0) -- (4, 4);
  \node[red!60!white, opacity=0.9, right] at (4, 1) {$\Gamma_D$};

  \draw[blue!60!white] (0, 0) -- (4, 0);
  \draw[blue!60!white] (0, 4) -- (4, 4);
  \node[blue!60!white, opacity=0.9, below] at (3, 0) {$\Gamma_N$};

@@

Let $\Omega=\Omega^{+}\cup\Omega^{-}$. Introducing a Lagrange multiplier
$p=\epsilon^{-1}\jump{\nabla u \cdot n}\in Q=H^{-1/2}(\Gamma)$
the weak form of \eqref{eq:coupled} reads: Find $u\in V=H^1_{0, \Gamma_D}(\Omega)$,
$\hat{v}\in\hat{V}=H^1(\Gamma)$, $p\in Q$ such that
%
\begin{equation}\label{eq:coupled_weak}
  \begin{aligned}
    &\int_{\Omega}\nabla u\cdot \nabla v &\phantom{+\int_{\Gamma}\nabla \hat{u}\cdot \nabla \hat{v}} &+\epsilon\int_{\Gamma} p v &= \int_{\Omega}f v \quad &\forall v\in V,\\
    &\phantom{\int_{\Omega}\nabla u\cdot \nabla v} &+{\int_{\Gamma}\nabla \hat{u}\cdot \nabla \hat{v} + \hat{u}\hat{v}} &-\int_{\Gamma} p\hat{v} &= \int_{\Gamma}\hat{f}\hat{v} \quad &\forall \hat{v}\in \hat{V},\\
    &\epsilon\int_{\Gamma}u q &-\int_{\Gamma}\hat{u}q &\phantom{-\int{\Gamma}q\hat{v}} &= \int_{\Gamma}h q \quad &\forall q\in Q.
  \end{aligned}
\end{equation}

In the following it will be convenient to work with the operator form of
\eqref{eq:coupled_weak}. In particular the left-hand-side defines an operator
%
\begin{equation}\label{eq:operator2d}
  \mathcal{A} = \begin{pmatrix}
    -\Delta_{\Omega} & & -\epsilon T^{\prime}\\
    & -\Delta_{\Gamma} + I & I\\
    -\epsilon T & I & \\
  \end{pmatrix} =
  \begin{pmatrix}
    A_2 &  & B_2'\\
    & A_1 & -B_1'\\
    B_2 & -B_1 & \\
   \end{pmatrix}.
\end{equation}
In \cite{kuchta2016preconditioners} it is shown that $\mathcal{A}$ is an
isomorphism from $W$ to its dual space where
\[
W=H^1(\Omega)_{0, \Gamma_D}\times H^1{(\Gamma)}\times(\epsilon H^{-1/2}(\Gamma)\cap H^{-1}(\Gamma)).
\]
It is further shown that discretization in terms of continuous linear Lagrange
elements (P1) for all the unknowns is inf-sup stable and in turn the diagonal operator
\begin{equation}\label{eq:P1_precond}
\mathcal{B}=\text{diag}(-\Delta_{\Omega}, -\Delta_{\Gamma}+I, \epsilon^2(-\Delta_{\Gamma})^{-1/2}+(-\Delta_{\Gamma}+I)^{-1})^{-1}
\end{equation}
is a robust preconditioner for $\mathcal{A}$. For the sake of completness
(and verifying the implementation) Table \ref{tab:P1} shows the number of
MinRes iterations and condition numbers of the preconditioned problem for
different levels of refinements and values of the parameter $\epsilon$. The
stability is evident.
%
\begin{table}
    \footnotesize{
      \begin{minipage}{0.49\textwidth}
          \begin{center}
            \begin{tabular}{c|cccc} \hline
              \multirow{2}{*}{$\epsilon$} & \multicolumn{4}{c}{$h$}\\
              \cline{2-5}
 &   $2^{-2}$  & $2^{-3}$ & $2^{-4}$ & $2^{-5}$   \\  
\hline
$10^{6}$ &	7.73&	7.82&	7.86&	7.87\\
$10^{4}$ &	7.73&	7.82&	7.86&	7.87\\
$10^{2}$   &	7.73&	7.82&	7.86&	7.87\\
1     &	7.39&	7.79&	8.01&	8.13\\
$10^{-2}$  &	2.62&	2.63&	2.63&	2.65\\
$10^{-4}$ &	2.62&	2.62&	2.62&	2.62\\
$10^{-6}$ &	2.62&	2.62&	2.62&	2.62\\
\hline
\end{tabular}
\end{center}
      \end{minipage}
    }
          \footnotesize{
            \begin{minipage}{0.49\textwidth}
              \begin{center}
            \begin{tabular}{c|cccccc} \hline
              \multirow{2}{*}{$\epsilon$} & \multicolumn{6}{c}{$h$}\\
              \cline{2-7}
 &   $2^{-3}$ & $2^{-4}$ & $2^{-5}$ & $2^{-6}$ & $2^{-7}$ & $2^{-8}$  \\  
              \hline
$10^{6}$ & 21 & 22 & 22 & 21 & 21 & 19\\
$10^{4}$ & 21 & 22 & 22 & 21 & 21 & 19\\
$10^{2}$ & 21 & 22 & 22 & 21 & 21 & 19\\
1 & 24 & 28 & 26 & 26 & 27 & 25\\
$10^{-2}$ & 10 & 10 & 10 & 12 & 13 & 14\\
$10^{-4}$ & 6 & 6 & 6 & 6 & 7 & 7\\
$10^{-6}$ & 4 & 4 & 4 & 4 & 4 & 4\\
\hline
\end{tabular}                
              \end{center}
        \end{minipage}
      }
          \caption{
            Problem \eqref{eq:coupled_weak} discretized with P1 elements.
            Condition number (left) and iterations counts (right) using
            preconditioner \eqref{eq:P1_precond}.
          }
          \label{tab:P1} 
%\label{table1}
\end{table}

@@

  \includegraphics[width=0.4\textwidth]{./img/conform_fine.tex}\hfill
  \includegraphics[width=0.4\textwidth]{./img/cut_fine.tex}\\
  \includegraphics[width=0.4\textwidth]{./img/nocut_lm2d.tex}\hfill
    \includegraphics[width=0.4\textwidth]{./img/cut_lm2d.tex}
  \end{center}
  \caption{Schematic geometric setting and discretization of mutliscale problems.
    Position of interface $\Gamma$ is denoted by \figlegend{green!50!black}. Meshes
    underlying discrete approximation spaces $\hat{V}_h$ and $Q_h$ of $\hat{V}$ and $Q$ are denoted respectively
    by \figlegend{white!50!blue} and \figlegend{white!50!red}. Geometric correspondence
    is signified by dashed line. Cases: (i) trace mesh of $V_h$ is used for $\hat{V}_h$ and $Q_h$,
    (ii) $\Gamma$ does not intersect $\Omega_h $ cell interiors, (iii) $\Gamma$ does
    intersect cell interiors, (iv) as (ii) with $Q_h$ defined on intersected cells and
    (v) as (iii) with $Q_h$ defined on intersected cells.
  }
 \label{fig:domains}
\end{figure}

The rest of the report is structured as follows. Considering \eqref{eq:operator2d}
we first treat the problems related to domains $\Omega$, i.e. $\mathcal{A}_2=\left[A_2, B_2'; B_2, 0\right]$,
and $\Gamma$, i.e. $\mathcal{A}_1=\left[A_1, B_1'; B_1, 0\right]$, individually\footnote{This is done for
clarity of exposition, to better fix the ideas and debugging}. More precisely, in
\S\ref{sec:problem_omega} a set of preconditioners for different formulations of the Babu{\v s}ka
problem is presented culminating in a formulation having the discrete Lagrange multiplier on the
cut cells. In \S\ref{sec:problem_gamma} we deal with instabities in $\mathcal{A}_1$ which arise
due to choices of discretization of the Lagrange multiplier space in $\mathcal{A}_2$. Finally in
\S\ref{sec:coupled} the insights from individual problems are combined to yield robust solvers
for the coupled problem.

\section{Coupling 2$d$-1$d$}\label{sec:problem_omega}
\input{problem_omega}

\section{Coupling 1$d$-1$d$}\label{sec:problem_gamma}
\input{problem_gamma}

\section{Coupling 2$d$-1$d$-1$d$}\label{sec:coupled}
\input{coupled}

>>>>

\newcommand{\MK}[1]{\textcolor{cyan}{#1}}

\newcommand{\semi}[1]{\lvert{#1}\rvert}
\newcommand{\norm}[1]{\lVert{#1}\rVert}
\newcommand{\jump}[1]{\ensuremath{[\![#1]\!]} }
\newcommand{\avg}[1]{\ensuremath{\left\{\!\left\{#1\right\}\!\right\}} }

\def\ud{u_{\odot}}
\def\vd{v_{\odot}}
\def\ld{\lambda_{\odot}}
\def\md{\mu_{\odot}}
\def\lld{l_{\odot}}
\def\uf{u_{\ominus}}
\def\up{u_{\oplus}}
\def\eps{\epsilon}
\def\nn{\boldsymbol n}
\def\rr{\boldsymbol r}
\def\RR{\boldsymbol R}
\def\kk{\boldsymbol k}
\def\ss{\boldsymbol s}
\def\uu{\boldsymbol u}
\def\vv{\boldsymbol v}
\def\xx{\boldsymbol x}
\def\bu{\overline{u}}
\def\bv{\overline{v}}
\def\tu{\widetilde{u}}
\def\tv{\widetilde{v}}
\def\TT{\boldsymbol T}
\def\NN{\boldsymbol N}
\def\BB{\boldsymbol B}
\def\ttu{\widetilde{\widetilde{u}}}
\def\ttv{\widetilde{\widetilde{v}}}
\def\cv{\check{v}}
\def\mesh{{\cal T}^h}
\def\ball{{\cal B}}
\def\R{\mathbb{R}}
\def\D{\mathcal{D}}
\def\DD{\partial\mathcal{D}}
\def\trace{{\mathcal{T}_\Gamma}}
\def\mtrace{{\overline{\mathcal{T}}_\Lambda}}
\def\ext{\mathcal{E}_\Gamma}
\def\ide{\mathcal{I}}
\def\ii{\hat{\imath}}
\newcommand{\avrd}[1]{\overline{\overline{#1}}}
\newcommand{\avrc}[1]{\overline{#1}}
\newcommand{\refe}[1]{{#1}_{\mathrm{ref}}}

@@

  % Boundaries
  \draw[very thick, red!60!white] (0, 0) -- (0, 4);
  \draw[very thick,red!60!white] (4, 0) -- (4, 4);
  \node[red!60!white, opacity=0.9, right] at (4, 1) {$\Gamma_D$};

  \draw[very thick, blue!60!white] (0, 0) -- (4, 0);
  \draw[very thick, blue!60!white] (0, 4) -- (4, 4);
  \node[blue!60!white, opacity=0.9, below] at (3, 0) {$\Gamma_N$};

@@

Let $\Omega=\Omega^{+}\cup\Omega^{-}$. Introducing a Lagrange multiplier
$p=\epsilon^{-1}\jump{\nabla u \cdot n}\in Q=H^{-1/2}(\Gamma)$
the weak form of \eqref{eq:coupled} reads: Find $u\in V=H^1_{0, \Gamma_D}(\Omega)$,
$\hat{v}\in\hat{V}=H^1(\Gamma)$, $p\in Q$ such that
%
\begin{equation}\label{eq:coupled_weak}
  \begin{aligned}
    &\int_{\Omega}\nabla u\cdot \nabla v\,\mathrm{d}x &\phantom{+\int_{\Gamma}\nabla \hat{u}\cdot \nabla \hat{v}} &+\epsilon\int_{\Gamma} p v\,\mathrm{d}s &= \int_{\Omega}f v\,\mathrm{d}x \quad &\forall v\in V,\\
    &\phantom{\int_{\Omega}\nabla u\cdot \nabla v} &+{\int_{\Gamma}\nabla \hat{u}\cdot \nabla \hat{v} + \hat{u}\hat{v}\,\mathrm{d}x} &-\int_{\Gamma} p\hat{v}\,\mathrm{d}s &= \int_{\Gamma}\hat{f}\hat{v}\,\mathrm{d}s \quad &\forall \hat{v}\in \hat{V},\\
    &\epsilon\int_{\Gamma}u q\,\mathrm{d}s &-\int_{\Gamma}\hat{u}q\,\mathrm{d}s &\phantom{-\int{\Gamma}q\hat{v}} &= \int_{\Gamma}h q\,\mathrm{d}s \quad &\forall q\in Q.
  \end{aligned}
\end{equation}

In the following it will be convenient to work with the operator form of
\eqref{eq:coupled_weak}. In particular the left-hand-side defines an operator
%
\begin{equation}\label{eq:operator2d}
  \mathcal{A} = \begin{pmatrix}
    -\Delta_{\Omega} & & -\epsilon T^{\prime}\\
    & -\Delta_{\Gamma} + I & -I\\
    -\epsilon T & -I & \\
  \end{pmatrix} =
  \begin{pmatrix}
    A_2 &  & B_2'\\
    & A_1 & -B_1'\\
    B_2 & -B_1 & \\
   \end{pmatrix}.
\end{equation}
In \cite{kuchta2016preconditioners} it is shown that $\mathcal{A}$ is an
isomorphism from $W$ to its dual space where
\[
W=H^1(\Omega)_{0, \Gamma_D}\times H^1{(\Gamma)}\times(\epsilon H^{-1/2}(\Gamma)\cap H^{-1}(\Gamma)).
\]
It is further shown that discretization in terms of continuous linear Lagrange
elements (P1) for all the unknowns is inf-sup stable and in turn the diagonal operator
\begin{equation}\label{eq:P1_precond}
\mathcal{B}=\text{diag}(-\Delta_{\Omega}, -\Delta_{\Gamma}+I, \epsilon^2(-\Delta_{\Gamma})^{-1/2}+(-\Delta_{\Gamma}+I)^{-1})^{-1}
\end{equation}
is a robust preconditioner for $\mathcal{A}$. For the sake of completness
(and verifying the implementation) Table \ref{tab:P1} shows the number of
MinRes iterations and condition numbers\footnote{
  Unless specified otherwise all the experiments are done on $\Omega$ a unit square
  triangulated in $2N^2$ isosceles triangles. MinRes iterations are always started
  from a random initial vector and terminate one the relative preconditioned residual
  norm is reduced by factor $10^{8}$. Condition numbers of the preconditioned problem
  are computed by direct solvers, i.e. from the full spectrum.
} of the preconditioned problem for
different levels of refinement and values of the parameter $\epsilon$. The
stability is evident.
%
\begin{table}
    \footnotesize{
      \begin{minipage}{0.49\textwidth}
          \begin{center}
            \begin{tabular}{c|cccc} \hline
              \multirow{2}{*}{$\epsilon$} & \multicolumn{4}{c}{$h$}\\
              \cline{2-5}
 &   $2^{-2}$  & $2^{-3}$ & $2^{-4}$ & $2^{-5}$   \\  
\hline
$10^{6}$ &	7.73&	7.82&	7.86&	7.87\\
$10^{4}$ &	7.73&	7.82&	7.86&	7.87\\
$10^{2}$   &	7.73&	7.82&	7.86&	7.87\\
1     &	7.39&	7.79&	8.01&	8.13\\
$10^{-2}$  &	2.62&	2.63&	2.63&	2.65\\
$10^{-4}$ &	2.62&	2.62&	2.62&	2.62\\
$10^{-6}$ &	2.62&	2.62&	2.62&	2.62\\
\hline
\end{tabular}
\end{center}
      \end{minipage}
    }
          \footnotesize{
            \begin{minipage}{0.49\textwidth}
              \begin{center}
            \begin{tabular}{c|cccccc} \hline
              \multirow{2}{*}{$\epsilon$} & \multicolumn{6}{c}{$h$}\\
              \cline{2-7}
 &   $2^{-3}$ & $2^{-4}$ & $2^{-5}$ & $2^{-6}$ & $2^{-7}$ & $2^{-8}$  \\  
              \hline
$10^{6}$ & 21 & 22 & 22 & 21 & 21 & 19\\
$10^{4}$ & 21 & 22 & 22 & 21 & 21 & 19\\
$10^{2}$ & 21 & 22 & 22 & 21 & 21 & 19\\
1 & 24 & 28 & 26 & 26 & 27 & 25\\
$10^{-2}$ & 10 & 10 & 10 & 12 & 13 & 14\\
$10^{-4}$ & 6 & 6 & 6 & 6 & 7 & 7\\
$10^{-6}$ & 4 & 4 & 4 & 4 & 4 & 4\\
\hline
\end{tabular}                
              \end{center}
        \end{minipage}
      }
          \caption{
            Problem \eqref{eq:coupled_weak} discretized with P1 elements.
            Condition number (left) and iterations counts (right) using
            preconditioner \eqref{eq:P1_precond}.
          }
          \label{tab:P1} 
%\label{table1}
\end{table}

@@

  \includegraphics[width=0.4\textwidth]{./img/conform_fine.tex}\hfill
  \includegraphics[width=0.4\textwidth]{./img/cut_fine.tex}\\
  \includegraphics[width=0.4\textwidth]{./img/nocut_lm2d.tex}\hfill
    \includegraphics[width=0.4\textwidth]{./img/cut_lm2d.tex}
  \end{center}
  \caption{Schematic geometric setting and discretization of multiscale problems.
    Position of interface $\Gamma$ is denoted by \figlegend{green!50!black}. Meshes
    underlying discrete approximation spaces $\hat{V}_h$ and $Q_h$ of $\hat{V}$ and $Q$ are denoted respectively
    by \figlegend{white!50!blue} and \figlegend{white!50!red}. Geometric correspondence
    is signified by dashed line. Cases: (i) trace mesh of $V_h$ is used for $\hat{V}_h$ and $Q_h$,
    (ii) $\Gamma$ does not intersect $\Omega_h $ cell interiors, (iii) $\Gamma$ does
    intersect cell interiors, (iv) as (ii) with $Q_h$ defined on intersected cells and
    (v) as (iii) with $Q_h$ defined on intersected cells.
  }
 \label{fig:domains}
\end{figure}

The rest of the report is structured as follows. Considering \eqref{eq:operator2d}
we first treat the problems related to domains $\Omega$, i.e. $\mathcal{A}_2=\left[A_2, B_2'; B_2, 0\right]$,
and $\Gamma$, i.e. $\mathcal{A}_1=\left[A_1, B_1'; B_1, 0\right]$, individually\footnote{This is done for
clarity of exposition, to better fix the ideas and debugging.}. More precisely, in
\S\ref{sec:problem_omega} a set of preconditioners for different formulations of the Babu{\v s}ka
problem is presented culminating in a formulation having the discrete Lagrange multiplier on the
cut cells. In \S\ref{sec:problem_gamma} we deal with instabities in $\mathcal{A}_1$ which arise
due to choices of discretization of the Lagrange multiplier space in $\mathcal{A}_2$. Finally in
\S\ref{sec:coupled} the insights from individual problems are combined to yield robust solvers
for the coupled problem.

\section{Preconditioning $\mathcal{A}_2$}\label{sec:problem_omega}
\input{problem_omega}

\section{Preconditioning $\mathcal{A}_1$}\label{sec:problem_gamma}
\input{problem_gamma}

\section{Preconditioning coupled problem}\label{sec:coupled}
\input{coupled}
